Re-analysing the data from @moffatt_inner_2020: What can we learn from an under-powered absence of difference?

### 2022-02-11 manuscript.Rmd
manucript second revision

---
title: "Re-analysing the data from @moffatt_inner_2020: What can we learn from an under-powered absence of difference?"
shorttitle: "The absence of evidence fallacy"
# note: \textcolor{red}{This is an example of a note}
author: 
  - name: "Ladislas Nalborczyk"
    affiliation: "1, 2, 3"
    corresponding: yes
    address: "LPC, CNRS, Aix-Marseille University, 13 Boulevard Gustave Desplaces, 13003 Marseille, France"
    email: "ladislas.nalborczyk@univ-amu.fr"
affiliation:
  - id: 1
    institution: Univ. Grenoble Alpes, CNRS, Grenoble INP, GIPSA-lab, 38000 Grenoble, France
  - id: 2
    institution: Aix Marseille Univ, CNRS, LPC, Marseille, France
  - id: 3
    institution: Aix Marseille Univ, CNRS, LNC, Marseille, France
abstract: "@moffatt_inner_2020 reported the results of an experiment (N = 26 in the final sample) comparing the facial electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference in the perioral muscular activity between the rumination and distraction conditions, @moffatt_inner_2020 concluded that *self-reported* inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. We suggest this conclusion is at best hasty. Indeed, concluding on the absence of an effect based on an under-powered non-significant *p*-value is strongly uninformative. In this short commentary, we show that there is limited evidence for the main conclusion put forward by @moffatt_inner_2020. We suggest ways forward, both from a theoretical and methodological perspective. Complete source code, reproducible analyses, and figures are available at https://osf.io/ba3gk/."
keywords: NHST, Bayes factor, reanalysis, inner speech, rumination, electromyography
csl: csl/apa7.csl
classoption: man, donotrepeattitle
linenumbers: yes
floatsintext: yes
link-citations: true
numbersections: false
mask: yes # masking author names
bibliography: [bib/library.bib, bib/r-references.bib]
nocite: | 
  @R-base, @R-knitr, @R-rmarkdown, @R-tidyverse, @R-papaja,
  @R-wordcountaddin, @R-here
output:
  papaja::apa6_pdf:
    keep_tex: true
    highlight: tango
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(wordcountaddin)
library(BayesFactor)
library(ggbeeswarm)
library(rmarkdown)
library(tidyverse)
library(patchwork)
library(parallel)
library(papaja)
library(readxl)
library(knitr)
library(BEST)
library(brms)
library(glue)
library(here)
library(pwr)

@@

# counting words in the current .Rmd file
wordcount <- word_count(knitr::current_input() )
```

Wordcount (excluding abstract, references, tables, and figures): `r wordcount`

\newpage

@@

Given the predominantly verbal nature of rumination [e.g., @ehring_repetitive_2008; @goldwin_concreteness_2012; @goldwin_concreteness_2013; @mclaughlin_effects_2007], we previously proposed to consider rumination as a form of inner speech and to study it using the methods that have historically been used to study other forms of inner speech, namely, by using surface electromyography (EMG) and motor interference protocols [e.g., @nalborczyk_dissociating_2021; @nalborczyk_orofacial_2017; @nalborczyk_understanding_2019; @nalborczyk_articulatory_2020]. We first showed that induced rumination was accompanied by increased facial (both over a forehead and a perioral site) muscular activity in comparison to a rest period [@nalborczyk_orofacial_2017]. However, because rumination was only compared to a rest period, it remained uncertain whether this perioral activity was specifically related to (inner) speech processes. Therefore, we ran a follow-up study comparing verbal to non-verbal rumination, which suggested that the facial EMG correlates we had previously identified were not specifically related to the verbal content of the ruminative thoughts [@nalborczyk_dissociating_2021]. We discussed these findings in length and proposed several theoretical interpretations that can account for these results in the discussion of @nalborczyk_dissociating_2021 and more extensively in @nalborczyk_understanding_2019.

Although these points were not addressed by @moffatt_inner_2020, their experimental design nevertheless had the potential to inform our understanding of the involvement of the speech motor system in different varieties of inner speech as well as to clarify the relation between the peripheral correlates of inner speech and (self-reported) subjective experience.

@moffatt_inner_2020 designed an experiment with the aim of refining our understanding of the involvement of the speech motor system in different varieties of inner speech and clarifying the relation between the peripheral correlates of inner speech and (self-reported) subjective experience. Their main conclusion is that inner experience between induced rumination and distraction differs "without a change in electromyographic correlates of inner speech" (p.1). In other words, they suggest that the subjective experience of inner speech is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented mostly by the EMG amplitude recorded over the orbicularis oris inferior and orbicularis oris superior muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative for other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference should be substantiated by adequately-powered statistical tests (given the target effect size) as well as reliable measures. This is unlikely to be the case here, for reasons that we will present and discuss in the present article.

Moreover, a simple visual exploration of the data reveals important variability between individuals in the main effect of interest. That is, some participants had higher perioral muscular activity in the rumination condition than in the distraction condition, and some other participants showed the reverse pattern. This suggests unexplored variation in the determinants of this effect (e.g., the content of the inner experience). In the following, we explore each of these limitations and suggests ways forward, both from a theoretical and from a methodological perspective.

Indeed, the relation between the inner experience and the physiological correlates of inner speech production was only inferred from group averages. However, given the important inter-individual variability, this reasoning appears highly problematic

# Re-analysing the data from @moffatt_inner_2020

@@

To model EMG peak amplitude variations in response to the rumination and distraction inductions, we fitted a Bayesian multivariate regression model with the natural logarithm of the EMG peak amplitude as an outcome and *Condition* (baseline, rumination, distraction) as a categorical predictor. Therefore, the intercept represents the estimated natural logarithm of the EMG peak amplitude in the baseline condition, and the slopes for the rumination and distraction conditions represent deviations from the baseline. These analyses were conducted using the `brms` package [@R-brms], an `R` implementation of Bayesian multilevel models that employs the probabilistic programming language `Stan` [@carpenter_stan_2017]. We ran four chains including each 10.000 iterations and a warmup of 2.000 iterations, using weakly informative priors on all parameters (for more details, see the [supplementary materials](#supp)). Posterior convergence was assessed examining autocorrelation and trace plots, as well as the Gelman-Rubin statistic. Constant effects estimates were summarised via their posterior mean and 95% credible interval. We also report Bayes factors (BFs) computed using the Savage-Dickey method.^[This method consists in taking the ratio of the posterior density at the point of interest divided by the prior density at that point [@wagenmakers_bayesian_2010].] These BFs can be interpreted as updating factors, from prior knowledge (what we knew before seeing the data) to posterior knowledge (what we know after seeing the data). A summary of the estimations from this model is presented in Table \@ref(tab:summary). This analysis revealed strong evidence for the hypothesis of a higher average EMG peak amplitude in the rumination condition as compared to the baseline condition for both the FRO and OOI muscles (as assessed by the BFs). However, the BFs supported the null hypothesis (i.e., the hypothesis of no difference) between the baseline and distraction conditions for the FRO and were inconclusive for both the OOI and OOS muscles.

```{r models, results = "hide"}
##############################
# Bayesian multilevel models #
##############################

# priors for the constant-effects model
priors_constant <- c(
    prior(normal(0, 5), class = b, coef = Intercept, resp = "FRO"),
    prior(normal(0, 1), class = b, resp = "FRO"),
    prior(exponential(1), class = sigma, resp = "FRO"),
    prior(normal(0, 5), class = b, coef = Intercept, resp = "OOS"),
    prior(normal(0, 1), class = b, resp = "OOS"),
    prior(exponential(1), class = sigma, resp = "OOS"),
    prior(normal(0, 5), class = b, coef = Intercept, resp = "OOI"),
    prior(normal(0, 1), class = b, resp = "OOI"),
    prior(exponential(1), class = sigma, resp = "OOI")
    )

# fitting the model
constant_effects <- brm(
    mvbind(FRO, OOS, OOI) ~ 0 + Intercept + condition,
    family = gaussian(),
    prior = priors_constant,
    data = df2,
    chains = 4, cores = 4,
    warmup = 2000, iter = 10000,
    control = list(adapt_delta = 0.95),
    sample_prior = TRUE,
    file = here("models/constant_effects")
    )

# priors for the varying-effects model
priors_varying <- c(
    prior(normal(0, 5), class = b, coef = Intercept, resp = "FRO"),
    prior(normal(0, 1), class = b, resp = "FRO"),
    prior(exponential(1), class = sigma, resp = "FRO"),
    prior(exponential(1), class = sd, resp = "FRO"),
    prior(normal(0, 5), class = b, coef = Intercept, resp = "OOS"),
    prior(normal(0, 1), class = b, resp = "OOS"),
    prior(exponential(1), class = sigma, resp = "OOS"),
    prior(exponential(1), class = sd, resp = "OOS"),
    prior(normal(0, 5), class = b, coef = Intercept, resp = "OOI"),
    prior(normal(0, 1), class = b, resp = "OOI"),
    prior(exponential(1), class = sigma, resp = "OOI"),
    prior(exponential(1), class = sd, resp = "OOI")
    )

# fitting the model
varying_effects <- brm(
    mvbind(FRO, OOS, OOI) ~ 0 + Intercept + condition + (0 + Intercept | ID),
    family = gaussian(),
    prior = priors_varying,
    data = df2,
    chains = 4, cores = 4,
    warmup = 2000, iter = 10000,
    control = list(adapt_delta = 0.95),
    sample_prior = TRUE,
    file = here("models/varying_effects")
    )
```

```{r bf, eval = FALSE, fig.width = 10, fig.cap = "Savage-Dickey Bayes factor for the difference between the rumination and distraction conditions for each muscle. The BF is computed as the ratio of the posterior density to the prior density at $\\theta = 0$."}
# BF testing
hyp_fro <- hypothesis(
    x = varying_effects,
    hypothesis = "FRO_conditionDIS = FRO_conditionRUM"
    )

p1 <- plot(hyp_fro, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5), ylim = c(0, 7.5) ) +
    theme_bw(base_size = 12, base_family = "Open Sans") +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      y = "Probability density",
      title = paste0(
        "BF10 = ", round(1 / hyp_fro$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_fro$hypothesis$Evid.Ratio, 3), ")"
        )
      )

hyp_ooi <- hypothesis(
    varying_effects,
    "OOI_conditionDIS = OOI_conditionRUM"
    )

p2 <- plot(hyp_ooi, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5), ylim = c(0, 7.5) ) +
    theme_bw(base_size = 12, base_family = "Open Sans") +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      title = paste0(
        "BF10 = ", round(1 / hyp_ooi$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_ooi$hypothesis$Evid.Ratio, 3), ")"
        )
      )

hyp_oos <- hypothesis(
    varying_effects,
    "OOS_conditionDIS = OOS_conditionRUM"
    )

p3 <- plot(hyp_oos, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5), ylim = c(0, 7.5) ) +
    theme_bw(base_size = 12, base_family = "Open Sans") +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      title = paste0(
        "BF10 = ", round(1 / hyp_oos$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_oos$hypothesis$Evid.Ratio, 3), ")"
        )
      )

p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

```{r summary, result = "asis"}
changeSciNot <- function(n) {

  output <- format(n, digits = 4, scientific = TRUE)
  output <- sub("e", "*10^", output)
  output <- sub("\\+0?", "", output)
  output <- sub("-0?", "-", output)
  output

summary_varying_effects <-
    # retrieve estimations from bmod5
    summary(varying_effects, prob = 0.95)$fixed %>%
    # convert to dataframe
    data.frame %>%
    # sets rownames as the first column
    rownames_to_column(var = "response") %>%
    # removing ESS information
    dplyr::select(-Bulk_ESS, -Tail_ESS) %>%
    # rename columns
    magrittr::set_colnames(
        c("Term", "Estimate", "SE", "Lower", "Upper", "Rhat")
        ) %>%
    # compute BF for each effect
    mutate(
        BF10 = 1 / hypothesis(
          varying_effects, glue("{Term} = 0")
          )$hypothesis$Evid.Ratio
        ) %>%
    # round BFs
    mutate(
      BF10 = ifelse(
        BF10 < 0.001 | BF10 > 1e3,
        changeSciNot(abs(BF10) ),
        round(BF10, 3)
        )
      ) %>%
    # if numeric, round to 3 decimals
    mutate_if(is.numeric, round, 3)

apa_table(
    summary_varying_effects,
    placement = "H",
    digits = 3,
    align = c("l", rep("c", 6) ),
    caption = "Estimated value of the natural logarithm of the EMG peak amplitude in each condition and for each muscle.",
    note = "For each effect, the 'Estimate' reports the estimated average value of the natural logarithm of the EMG peak amplitude, followed by its standard error (SE). The 'Lower' and 'Upper' columns contain the lower and upper bounds of the 95% CrI, whereas the 'Rhat' column reports the Gelman-Rubin statistic. The last column reports the BF in favour of the alternative hypothesis (relative to the null hypothesis).",
    small = TRUE,
    escape = TRUE
    )
```

Because the result of a Bayesian analysis is a joint posterior probability over all parameters of the model, we can compute the posterior distribution of the difference between any pair of conditions. In Figure \@ref(fig:posterior), we represent the posterior distribution of the difference in EMG peak amplitude between the rumination and distraction conditions for each muscle. This figure reveals that the most probable value for this difference was $\beta = 0.228$ (95% CrI [0.098, 0.357]) for the FRO muscle, $\beta = 0.081$ (95% CrI [-0.155, 0.324]) for the OOI muscle, and $\beta = 0.047$ (95% CrI [-0.167, 0.27]) for the OOS muscle. Moreover, comparing the posterior distribution to $\theta = 0$ reveals that there is a probability of 0.753 that the average peak EMG amplitude recorded over the OOI is higher in the rumination condition than in the distraction condition [given the model, the priors, and the data from @moffatt_inner_2020].

```{r posterior, fig.width = 10, fig.height = 4, eval = TRUE, fig.cap = "Posterior distribution of the difference in EMG peak amplitude between the rumination and distraction condition for each muscle, along with its mean and 95\\% credible interval."}
# extracts posterior samples
post <- posterior_samples(varying_effects)

# grid of plots
par(mfrow = c(1, 3) )

plotPost(
    post$b_FRO_conditionRUM - post$b_FRO_conditionDIS,
    compVal = 0, xlab = "Rumination - Distraction", main = "FRO"
    )

plotPost(
    post$b_OOI_conditionRUM - post$b_OOI_conditionDIS,
    compVal = 0, xlab = "Rumination - Distraction", main = "OOI"
    )

plotPost(
    post$b_OOS_conditionRUM - post$b_OOS_conditionDIS,
    compVal = 0, xlab = "Rumination - Distraction", main = "OOS"
    )
```

@@

## Enthusiastic prior specifications

@@

```{r bf-dance, fig.cap = "Illustrating the distribution of Bayes factors in favour of the alternative hypothesis for different population effect sizes (N = 26). In the left panel, the effect size is fixed to d = 0 (i.e., the null hypothesis), in the middle panel, it is fixed to d = 0.36 (i.e., the supposed target effect size in Moffatt et al., 2020), and in the right panel, the effect size is fixed to d = 0.72 (i.e., the effect size reported in Nalborczyk et al., 2017). The red vertical dashed line indicates the value of the BF computed for the OOI by Moffatt et al. (2020), on the log scale."}
# inspired from http://daniellakens.blogspot.com/2016/07/dance-of-bayes-factors.html
# population effect size
cohensd <- 0.36

@@

# initialises an empty array to store the results
results <- data.frame(
  BF10 = numeric(nsims),
  BF10_null = numeric(nsims),
  BF10_double = numeric(nsims)
  )

for (i in 1:nsims) { # for each simulated experiment

  # produces N simulated participants
  x <- rnorm(n = n, mean = 0, sd = 1)

  # produces N simulated participants when d = 0.36
  y <- rnorm(n = n, mean = 0 + cohensd, sd = 1)

  # produces N simulated participants when d = 0
  y_null <- rnorm(n = n, mean = 0, sd = 1)

  # produces N simulated participants when d = 0.72
  y_double <- rnorm(n = n, mean = 0 + cohensd * 2, sd = 1)

  # computes the BF10 when d = 0.36
  results$BF10[i] <- ttestBF(
    x = x, y = y, paired = TRUE, rscale = "medium"
    ) %>%
    data.frame %>%
    pull(bf)

  # computes the BF10 when d = 0
  results$BF10_null[i] <- ttestBF(
    x = x, y = y_null, paired = TRUE, rscale = "medium"
    ) %>%
    data.frame %>%
    pull(bf)

  # computes the BF10 when d = 0.72
  results$BF10_double[i] <- ttestBF(
    x = x, y = y_double, paired = TRUE, rscale = "medium"
    ) %>%
    data.frame %>%
    pull(bf)

@@

# computing the proportion of simulated BFs above 1
prop_d0 <- mean(results$BF10_null > 1)
prop_d0.36 <- mean(results$BF10 > 1)
prop_d0.72 <- mean(results$BF10_double > 1)

# plotting the results
results %>%
  pivot_longer(cols = BF10:BF10_double, names_to = "type") %>%
  mutate(
    type = factor(
      type,
      levels = c("BF10_null", "BF10", "BF10_double"),
      labels = c("Cohen's d = 0", "Cohen's d = 0.36", "Cohen's d = 0.72")
      )
    ) %>%
  ggplot(aes(x = log(value), colour = type, fill = type) ) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_vline(xintercept = log(0.278), linetype = 2, colour = "orangered") +
  geom_histogram(colour = "white", show.legend = FALSE, bins = 40) +
  theme_bw(base_size = 12, base_family = "Open Sans") +
  facet_wrap(~ type, scales = "free_x") +
  labs(x = "Natural logarithm of the BF10", y = "Number of simulations") +
  scale_colour_brewer(palette = "Dark2", direction = 1) +
  scale_fill_brewer(palette = "Dark2", direction = 1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1) ) )
```

As shown in Figure \@ref(fig:bf-dance), the distribution of log-BFs computed under each hypothesis reveals important inter-simulation variability. For instance, under the null hypothesis, `r round(prop_d0 * 100, 2)`% of the computed log-BFs are above 0 and hence support the alternative hypothesis (although the "true" effect size is $d = 0$). When the "true" effect size is of $d = 0.36$, `r round(100 - prop_d0.36 * 100, 2)`% of the log-BFs are below 0 and hence support the null hypothesis (although the true effect size is actually non-null). When the "true" effect size is of $d = 0.72$, `r round(100 - prop_d0.72 * 100, 2)`% of the log-BFs are still below 0. In other words, for small sample and effect sizes, BFs have high error rates. To assess the extent to which the BF computed for the OOI by @moffatt_inner_2020 (i.e., $\text{BF}_{10} = 0.278$) is "surprising" given or "compatible" with some effect size, we can compute the proportion of simulated BFs that are equal or more extreme than the one reported by the authors for each hypothesis, which is approximately equal to `r round(1 - mean(results$BF10_null > 0.278), 2)`, `r round(1 - mean(results$BF10 > 0.278), 2)`, and `r round(1 - mean(results$BF10_double > 0.278), 2)`, for the hypothesis of $d = 0$, $d = 0.36$, and $d = 0.72$, respectively. In other words, observing a $\text{BF}_{10} = 0.278$ is more compatible with the hypothesis of $d = 0$ (i.e., the null hypothesis) than with the hypothesis of $d = 0.36$ or $d = 0.72$, for this sample size.

Second, we conducted another simulation with the aim of assessing the relation between the sample size and the value of the BF. In the previous section, we fitted a multivariate Bayesian regression model with varying-intercepts by participant and weakly informative priors on the EMG data collected by @moffatt_inner_2020. Using this model, we i) generated new datasets from the posterior predictive distribution of this model and ii) we computed the BF in favour of the alternative hypothesis ($\text{BF}_{10}$) using the `BayesFactor` package [@R-BayesFactor]. We used a "medium" prior (i.e., $r = \sqrt{2}/2$) on the scale parameter of the Cauchy prior for the alternative hypothesis. We repeated this procedure for varying sample sizes from 20 to 200 participants (by increments of 10 participants) with $1000$ simulations (i.e., 1000 simulated datasets) for each sample size.

```{r simulated-power, eval = FALSE, fig.cap = "Average natural logarithm of the Bayes factor in favour of the alternative hypothesis, along with its standard error, computed over 1000 datasets of increasing size simulated from the posterior predictive distribution of the varying-intercept multivariate Bayesian regresion model, fitted on the data from Moffatt et al. (2020). A log-BF belows 0 represents evidence for the null hypothesis (relative to the alternative hypothesis) and a log-BF above 0 represents evidence for the alternative hypothesis (relative to the null hypothesis)."}
# running the simulations
# source("code/bf_simulation.R")

# loading the results
load(file = "results/overall_results.Rda")

# simulations parameters
nsims <- 1e3
sample_size <- seq.int(from = 20, to = 200, by = 10)

# plotting the results
overall_results %>%
    na.omit() %>%
    filter_all(any_vars(. != 0) ) %>%
    pivot_longer(cols = bf_fro:bf_oos, names_to = "bf_type") %>%
    mutate(value = log(value) ) %>%
    mutate(bf_type = factor(bf_type, labels = c("FRO", "OOI", "OOS") ) ) %>%
    group_by(nobs, bf_type) %>%
    summarise(
        across(
            .cols = value,
            .fns = list(mean = mean, se = ~sd(.x) / sqrt(nsims) )
            )
        ) %>%
    ungroup() %>% 
    ggplot(aes(x = nobs, y = value_mean, colour = bf_type, fill = bf_type) ) +
    geom_hline(yintercept = 0, lty = 3) +
    geom_ribbon(
        aes(
          x = nobs, ymin = value_mean - 1.96 * value_se,
          ymax = value_mean + 1.96 * value_se, colour = NULL
          ),
        alpha = 0.5, show.legend = FALSE
        ) + 
    geom_line(show.legend = FALSE) +
    geom_point(show.legend = FALSE) +
    facet_wrap(~bf_type, scales = "free") +
    scale_x_continuous(breaks = unique(sample_size)[c(TRUE, FALSE)]) +
    theme_bw(base_size = 10) +
    labs(x = "Number of participants", y = "Natural logarithm of the Bayes factor") +
    scale_colour_brewer(palette = "Dark2", direction = 1) +
    scale_fill_brewer(palette = "Dark2", direction = 1)
```

As shown in Figure \@ref(fig:simulated-power), the natural logarithm of the BF in favour of the alternative hypothesis is growing proportionally with sample size. More precisely, whereas BFs computed on small samples (i.e., below 80 participants) support the null hypothesis, BFs computed on larger samples support the alternative hypothesis for all three facial muscles. For instance, the average $\text{BF}_{10}$ computed for the OOI muscle with a sample size of 160 participants is of $\exp(2.18) \approx 8.85$, indicating that these data are approximately $8.85$ times more likely under the alternative hypothesis than under the null hypothesis. To sum up, this reveals that although at low sample sizes, the BF may provide (weak) evidence for the null hypothesis (relative to the alternative hypothesis), this pattern may very well reverse for higher sample sizes.

>>>>

---
title: "Re-analysing the data from @moffatt_inner_2020: What can we learn from an under-powered absence of difference?"
shorttitle: "The absence of evidence fallacy"
author: 
  - name: "Ladislas Nalborczyk"
    affiliation: "1, 2, 3"
    corresponding: yes
    address: "LPC, CNRS, Aix-Marseille University, 13 Boulevard Gustave Desplaces, 13003 Marseille, France"
    email: "ladislas.nalborczyk@univ-amu.fr"
affiliation:
  - id: 1
    institution: Univ. Grenoble Alpes, CNRS, Grenoble INP, GIPSA-lab, 38000 Grenoble, France
  - id: 2
    institution: Aix Marseille Univ, CNRS, LPC, Marseille, France
  - id: 3
    institution: Aix Marseille Univ, CNRS, LNC, Marseille, France
abstract: "@moffatt_inner_2020 reported the results of an experiment (N = 26 in the final sample) comparing the facial electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference (and BFs between 3.6 and 4.3) in the perioral muscular activity between the rumination and distraction conditions, @moffatt_inner_2020 concluded that *self-reported* inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. In this short commlentary, we show that there is limited evidence for the main conclusion put forward by @moffatt_inner_2020, and we suggest ways forward, both from a theoretical and a methodological perspective. Complete source code, reproducible analyses, and figures are available at https://osf.io/ba3gk/."
keywords: NHST, Bayes factor, reanalysis, inner speech, rumination, electromyography
csl: csl/apa7.csl
classoption: man, donotrepeattitle
linenumbers: yes
floatsintext: yes
link-citations: true
numbersections: false
wordcount: "`r wordcountaddin::word_count(knitr::current_input() )` (excluding abstract, references, tables, and figures)"
# mask: yes # masking author names
bibliography: [bib/library.bib, bib/r-references.bib]
nocite: | 
  @R-base, @R-knitr, @R-rmarkdown, @R-tidyverse, @R-papaja,
  @R-wordcountaddin, @R-here
output:
  papaja::apa6_pdf:
    keep_tex: true
    highlight: tango
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(wordcountaddin)
library(BayesFactor)
library(simhelpers)
library(rmarkdown)
library(tidyverse)
library(parallel)
library(papaja)
library(readxl)
library(future)
library(furrr)
library(knitr)
library(glue)
library(here)
library(pwr)

@@

# counting words in the current .Rmd file
# wordcount <- word_count(knitr::current_input() )
```

@@

Given the predominantly verbal nature of rumination [e.g., @ehring_repetitive_2008; @goldwin_concreteness_2012; @goldwin_concreteness_2013; @mclaughlin_effects_2007], we previously proposed to consider rumination as a form of inner speech and to study it using the methods that have historically been used to study other forms of inner speech, namely, by using surface electromyography (EMG) and motor interference protocols [e.g., @nalborczyk_dissociating_2021; @nalborczyk_orofacial_2017; @nalborczyk_understanding_2019; @nalborczyk_articulatory_2022]. We first showed that induced rumination was accompanied by increased facial (both over a forehead and a perioral site) muscular activity in comparison to a rest period [@nalborczyk_orofacial_2017]. However, because rumination was only compared to a rest period, it remained uncertain whether this perioral activity was specifically related to (inner) speech processes. Therefore, we ran a follow-up study comparing verbal to non-verbal rumination, which suggested that the facial EMG correlates we had previously identified were not specifically related to the verbal content of the ruminative thoughts [@nalborczyk_dissociating_2021]. We discussed these findings in length and proposed several theoretical interpretations that can account for these results in the discussion of @nalborczyk_dissociating_2021 and more extensively in @nalborczyk_understanding_2019.

@moffatt_inner_2020 designed an experiment with the aim of refining our understanding of the involvement of the speech motor system in different varieties of inner speech and clarifying the relation between the peripheral correlates of inner speech and (self-reported) subjective experience. Their main conclusion is that inner experience between induced rumination and distraction differs "without a change in electromyographic correlates of inner speech" (p.1). In other words, they suggest that the subjective experience of inner speech is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented by the EMG amplitude recorded over the orbicularis oris inferior and orbicularis oris superior muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative for other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference should be substantiated by adequately-powered statistical tests (given the target effect size) as well as reliable measures. This is unlikely to be the case here, for reasons that we will present and discuss in the present article.

# Reconsidering the data from @moffatt_inner_2020

@@

@@

## Optimistic a priori power analysis

@@

```{r bf-dance, fig.cap = "Illustrating the distribution of Bayes factors in favour of the alternative hypothesis for different population effect sizes (N = 26). In the left panel, the effect size is fixed to d = 0 (i.e., the null hypothesis), in the middle panel, it is fixed to d = 0.36 (i.e., the supposed target effect size in Moffatt et al., 2020), and in the right panel, the effect size is fixed to d = 0.72 (i.e., the effect size reported in Nalborczyk et al., 2017). The red vertical dashed line indicates the value of the BF computed for the OOI by Moffatt et al. (2020), on the logarithmic scale. The grey shaded area represents the conventional (but questionable) interval in which BFs are usually considered as inconclusive."}
# inspired from http://daniellakens.blogspot.com/2016/07/dance-of-bayes-factors.html

@@

# defining the combinations of parameters values to explore
params <- crossing(cohensd = c(0, 0.36, 0.72), iterations = 1:nsims) %>%
    mutate(seed = round(runif(n = 1, min = 0, max = 1) * 2^30) + 1:n() )

run_sim <- function(cohensd, iterations, seed = NULL) {

    # setting the seed
    if (!is.null(seed) ) set.seed(seed)

    # produces N simulated participants in condition 1
    x <- rnorm(n = n, mean = 0, sd = 1)

    # produces N simulated participants in condition 2
    y <- rnorm(n = n, mean = 0 + cohensd, sd = 1)

    # computing the BF
    results <- ttestBF(x = x, y = y, paired = TRUE, rscale = "medium") %>%
    data.frame %>%
    pull(bf)

    # returning the BF
    return(results)

@@

# choosing an appropriate plan from the future package
plan(multisession(workers = detectCores() ) )

# then, running the simulations
simulation_results <- evaluate_by_row(
    params = params, sim_function = run_sim,
    .progress = FALSE, system_time = FALSE
    )

# computing the proportion of "erroneous" BFs
prop_error_d0 <- simulation_results %>% filter(cohensd == 0) %>% pull(.results)
prop_error_d0 <- mean(prop_error_d0 > 3)
prop_error_d0.36 <- simulation_results %>% filter(cohensd == 0.36) %>% pull(.results)
prop_error_d0.36 <- mean(prop_error_d0.36 < 1 / 3)
prop_error_d0.72 <- simulation_results %>% filter(cohensd == 0.72) %>% pull(.results)
prop_error_d0.72 <- mean(prop_error_d0.72 < 1 / 3)

# computing the proportion of inconclusive BFs
prop_inconclusive_d0 <- simulation_results %>% filter(cohensd == 0) %>% pull(.results)
prop_inconclusive_d0 <- mean(prop_inconclusive_d0 > 1/3 & prop_inconclusive_d0 < 3)
prop_inconclusive_d0.36 <- simulation_results %>% filter(cohensd == 0.36) %>% pull(.results)
prop_inconclusive_d0.36 <- mean(prop_inconclusive_d0.36 > 1/3 & prop_inconclusive_d0.36 < 3)
prop_inconclusive_d0.72 <- simulation_results %>% filter(cohensd == 0.72) %>% pull(.results)
prop_inconclusive_d0.72 <- mean(prop_inconclusive_d0.72 > 1/3 & prop_inconclusive_d0.72 < 3)

# plotting the results
simulation_results %>%
  mutate(
    type = factor(
      x = cohensd,
      labels = c("Cohen's d = 0", "Cohen's d = 0.36", "Cohen's d = 0.72")
      )
    ) %>%
  rename(value = .results) %>%
  ggplot(aes(x = log(value), colour = type, fill = type) ) +
  geom_rect(
    data = data.frame(xmin = log(1 / 3), xmax = log(3), ymin = -Inf, ymax = Inf),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    color = NA,
    alpha = 0.25,
    inherit.aes = FALSE
    ) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_histogram(alpha = 0.5, show.legend = FALSE, bins = 60) +
  geom_vline(xintercept = log(0.278), linetype = 1, size = 0.5, colour = "orangered") +
  annotate(geom = "point", x = log(0.278), y = 0, colour = "orangered", size = 2) +
  theme_bw(base_size = 12, base_family = "Open Sans") +
  facet_wrap(~ type) + # , scales = "free_x") +
  coord_cartesian(xlim = c(NA, 8) ) +
  labs(x = "Natural logarithm of the BF10", y = "Number of simulations") +
  scale_colour_brewer(palette = "Dark2", direction = 1) +
  scale_fill_brewer(palette = "Dark2", direction = 1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1) ) )
```

As shown in Figure \@ref(fig:bf-dance), the distribution of log-BFs computed under each hypothesis reveals important inter-simulation variability. For instance, `r round(prop_inconclusive_d0 * 100, 2)`% of the computed log-BFs under the null hypothesis are "inconclusive" and `r round(prop_error_d0 * 100, 2)`% of the log-BFs support the alternative hypothesis (although the population effect size is $d = 0$). When the population effect size is of $d = 0.36$, `r round(prop_inconclusive_d0.36 * 100, 2)`% of the computed log-BFs are "inconclusive" and `r round(prop_error_d0.36 * 100, 2)`% of the log-BFs support the null hypothesis (although the population effect size is actually non-null). When the population effect size is of $d = 0.72$, `r round(prop_inconclusive_d0.72 * 100, 2)`% of the computed log-BFs are "inconclusive" and `r round(prop_error_d0.72 * 100, 2)`% of the log-BFs support the null hypothesis. In brief, this simulations shows that for small sample and effect sizes, BFs have non-negligible error rates [see also @schonbrodt_sequential_2017].^[It should be noted that, as stressed by @rouder_optional_2014, Bayes factors indicate the relative evidence for a hypothesis, conditional on some observed data. In other words, Bayesian updating is not conditional on some hypothetical truth. With this in mind, the present simulation aims at illustrating how the frequentist properties of BFs may be used to design more informative studies [see also @schonbrodt_bayes_2018], while acknowledging that proper long-term error rates *control* is not the realm of the Bayesian framework.]

The problems discussed above about the interpretation of under-powered non-significant results also apply to the test @moffatt_inner_2020 performed regarding the effect or the conditions' order. In @nalborczyk_dissociating_2021, we manipulated the modality of rumination (whether it is verbal or non-verbal) in a between-subject manner to avoid order effects and to avoid dissipating the effects of the negative mood induction. More precisely, we assumed that inducing rumination after a distraction condition in a within-subject manner would dissipate the effects of the mood induction and therefore reduce the impact of the rumination induction. In contrast to this approach, @moffatt_inner_2020 asked participants to ruminate and then distract themselves (or reciprocally), after an induced stressor (an induced failure). Anticipating again that the order of the within-subject conditions may be an issue, @moffatt_inner_2020 say:
